---
title: "Assignment 4"
author: "Maharaj Teertha Deb, 40227747"
date: "2024-02-27"
output: pdf_document
---

# Section 3.1

## Problem 1 :

Consider the islands vector discussed in this section.

### (a)
Compare the histograms that result when using breaks based on Sturges' and Scott's rules. Make this comparison on the log scale and on the original scale.

### (a) Answer:

```{r}
data(islands)

par(mfrow=c(2,2))
hist(islands, breaks = "Sturges", main = "Sturges")
hist(islands, breaks = "Scott", main = "Scott")
hist(log(islands), breaks = "Sturges", main = "Sturges on log scale")
hist(log(islands), breaks = "Scott", main = "Scott on log scale")
```

###(b)
Construct a normal QQ plot, and compare the result with the plots in Figure 3.13; which one is most similar, and what does this tell you about this data set?

### (b) Answer:

```{r}
qqnorm(islands)
qqline(islands, col = "red")
```

### (c)
Construct a boxplot for these data on the log scale as well as the original scale.

### (c) Answer:

```{r}
par(mfrow=c(1,2))
boxplot(islands, main = "Original scale")
boxplot(log(islands), main = "Log scale")
```

###(d) 
Construct a dot chart of the areas. Is a log transformation needed here?

### (d) Answer:

```{r}
par(mfrow=c(1,2))
dotchart(islands,main="Original scale")
dotchart(log(islands),main="Log scale")
```

###(e)
Which form of graphic do you think is most appropriate for displaying these data?

### (e) Answer:

```{r}
dotchart(sort(log(islands)), main = "Log scale, sorted")

```

I think that's a curve graph of radical function because it looks like as the graph of sqrt(x)

## Problem 3:

Consider the pressure data frame. There are two columns temperature and pressure

### (a)

Construct a scatterplot with pressure on the vertical axis and temperature on the horizontal axis. Are the variables related linearly or nonlinearly?

### (a) Answer:

```{r}
plot(pressure ~ temperature, data = pressure)# Calculate correlation coefficient


correlation <- cor(pressure$temperature, pressure$pressure)

# Check for linearity
if(abs(correlation) < 0.005) {
  cat("The variables are related linearly (correlation =", correlation, ").\n")
} else {
  cat("The variables are related nonlinearly (correlation =", correlation, ").\n")
}

```

### (b)

The graph of the following function passes through the plotted points reasonably well: $y = (0.168 + 0.007x)^{\frac{20}{3}}$. The differences between the pressure values predicted by the curve and the observed pressure values are called residuals. Here is a way to calculate them: $$ \text{residuals} <- with(\text{pressure}\ , \text{pressure} - (0.168 + 0.007 \times \text{temperature})^{\frac{20}{3}} $$ Construct a normal QQ plot of these residuals and decide whether they are normally distributed or whether they follow a skewed distribution.

### (b) Answer:

```{r}
# Calculate residuals
predicted_pressure <- (0.168 + 0.007 * pressure$temperature)^(20/3)
residuals <- pressure$pressure - predicted_pressure

# Construct a normal QQ plot
qqnorm(residuals)
qqline(residuals, col = 2)

# Assess normality
shapiro.test(residuals)
```

### (c)

Now, apply the power transformation $y^{\frac{3}{20}}$ to the pressure data values. Plot these transformed values against temperature. Is a linear or nonlinear relationship evident now?

### (c) Answer:

```{r}
# Apply the power transformation y^(3/20) to pressure data values
transformed_pressure <- pressure$pressure^(3/20)

# Plot transformed values against temperature
plot(pressure$temperature, transformed_pressure,
     xlab = "Temperature", ylab = "Transformed Pressure",
     main = "Transformed Pressure vs Temperature")


# Calculate correlation coefficient between transformed pressure and temperature
correlation_transformed <- cor(pressure$temperature, transformed_pressure)

# Check for linearity
if(abs(correlation_transformed) < 0.005) {
  cat("The relationship between transformed pressure and temperature appears to 
      be linear (correlation =", correlation_transformed, ").\n")
} else {
  cat("The relationship between transformed pressure and temperature appears to 
      be nonlinear (correlation =", correlation_transformed, ").\n")
}


```

### (d)

Calculate residuals for the difference between transformed pressure values and those predicted by the straight line. Obtain a normal QQ plot, and decide whether the residuals follow a normal distribution or not.

### (d) Answer:

```{r}
# Fit a linear regression model to transformed pressure values and temperature
lm_model_transformed <- lm(transformed_pressure ~ pressure$temperature)

# Predict transformed pressure values using the linear model
predicted_transformed_pressure <- predict(lm_model_transformed)

# Calculate residuals
residuals_transformed <- transformed_pressure - predicted_transformed_pressure

# Construct a normal QQ plot of residuals
qqnorm(residuals_transformed)
qqline(residuals_transformed, col = 2)

# Assess normality
shapiro.test(residuals_transformed)

```

# Section 3.3

## Problem 4

Re-do the plots of Figure 3.18, but this time apply the six graphics functions to a variable $X$ which is defined by $X = Z^2$, where $Z$ is a simulated standard normal random variable as in the example. Comment on the changes that you see in the plots. (The variable $X$ is an example of a chi-squared random variable on one degree of freedom.) Refer to the previous question.

### Answer:

```{r}

# Load necessary library
library(MASS)  # For mvrnorm function

# Generate simulated standard normal random variable Z
set.seed(123)  # for reproducibility
Z <- rnorm(1000)

# Define variable X = Z^2
X <- Z^2

# Plotting Figure 3.18 with variable X
par(mfrow = c(3, 2))  # Set up a 3x2 layout for plots

# Plot histogram
hist(X, main = "Histogram of X")

# Plot ECDF
plot(ecdf(X), main = "Empirical Cumulative Distribution Function of X")

# Plot QQ plot
qqnorm(X, main = "QQ Plot of X")
qqline(X)

# Plot boxplot
boxplot(X, main = "Boxplot of X")

# Plot scatterplot of X against itself
plot(X, X, xlab = "X", ylab = "X", main = "Scatterplot of X against itself")

# Plot density plot
plot(density(X), main = "Density Plot of X")
```

## Problem 5:

Construct another set of six plots as in Figure 3.18, but this time applied to the data in EuStockMarkets. (i.e., apply the code of Figure 3.18 to $Z$, where $Z = \text{EuStockMarkets}$). Comment on the results, and use the summary()} function to gain further insight. Which of the six plots are useful descriptors of this dataset, and which may be limited in their usefulness?

### Answer:

```{r}

# Now, apply the same set of plots to EuStockMarkets data
data("EuStockMarkets")

# Plotting Figure 3.18 with EuStockMarkets data
Z <- EuStockMarkets[, 1]  # Using the first column of EuStockMarkets data
X <- Z^2

par(mfrow = c(3, 2))  # Set up a 3x2 layout for plots

# Plot histogram
hist(X, main = "Histogram of X")

# Plot ECDF
plot(ecdf(X), main = "Empirical Cumulative Distribution Function of X")

# Plot QQ plot
qqnorm(X, main = "QQ Plot of X")
qqline(X)

# Plot boxplot
boxplot(X, main = "Boxplot of X")

# Plot scatterplot matrix
plot(X, X, xlab = "X", ylab = "X", main = "Scatterplot of X against itself")

# Plot density plot
plot(density(X), main = "Density Plot of X")

# Comment on the results and use summary() function to gain further insight
summary(X)
```

# Section 3.4:

## Problem : 4

Consider the pressure data frame again.

### (a)

Plot pressure against temperature, and use the following command to pass a curve through these data:

$$
\text{curve} \left( (0.168 + 0.007x)^{\frac{20}{3}}, \text{from} = 0, \text{to} = 400, \text{add} = \text{TRUE} \right)
$$

### (a) Answer:

```{r}
# Assuming 'pressure' is your data frame with columns 'temperature' and 'pressure'
# If not, replace 'pressure' with the name of your data frame

# Plot pressure against temperature
plot(pressure$temperature, pressure$pressure, 
     xlab = "Temperature", ylab = "Pressure",
     main = "Plot of Pressure against Temperature")

# Pass a curve through the data
curve((0.168 + 0.007*x)^(20/3), from = 0, to = 400, add = TRUE)
```

### (b)

Now, apply the power transformation $y^{\frac{3}{20}}$ to the pressure data values. Plot these transformed values against temperature. Is a linear or nonlinear relationship evident now? Use the \texttt{abline()} function to pass a straight line through the points. (You need an intercept and slope for this – see the previous part of this question to obtain appropriate values.)

### (b) Answer:

```{r}

# Apply power transformation to pressure data values
transformed_pressure <- pressure$pressure^(3/20)

# Plot transformed values against temperature
plot(pressure$temperature, transformed_pressure,
     xlab = "Temperature", ylab = "Transformed Pressure")

# Pass a straight line through the points
abline(lm(transformed_pressure ~ pressure$temperature), col = "red")
```

\###(c)

### (c) Answer:

Add a suitable title to the graph.

```{r}
# Apply power transformation to pressure data values
transformed_pressure <- pressure$pressure^(3/20)

# Plot transformed values against temperature with title
plot(pressure$temperature, transformed_pressure,
     xlab = "Temperature", ylab = "Transformed Pressure",
     main = "Plot of Transformed Pressure against Temperature")

```

### (d)

Re-do the above plots, but use the \texttt{mfrow()} function to display them in a $2 \times 1$ layout on the graphics page. Repeat once again using a $1 \times 2$ layout.

### (d) Answer:

```{r}
# Apply power transformation to pressure data values
transformed_pressure <- pressure$pressure^(3/20)

# Create a 2 x 1 layout
par(mfrow = c(2, 1))

# Plot 1: Transformed pressure against temperature
plot(pressure$temperature, transformed_pressure,
     xlab = "Temperature", ylab = "Transformed Pressure",
     main = "Plot of Transformed Pressure against Temperature")

# Add a straight line passing through the points
abline(lm(transformed_pressure ~ pressure$temperature), col = "red")

# Plot 2: Pressure against temperature with curve
plot(pressure$temperature, pressure$pressure,
     xlab = "Temperature", ylab = "Pressure",
     main = "Plot of Pressure against Temperature")

# Pass a curve through the data
curve((0.168 + 0.007*x)^(20/3), from = 0, to = 400, add = TRUE)
```
