---
title: "Assignment 2"
author: "Maharaj Teertha Deb, 40227747"
date: "2024-02-01"
output: pdf_document
---

 Assignment 2

# TextBook Problems:

## Question 1:

```{latex}
Create the vector:
[1] 0 0 0 1 1 1 1 0 0 0 1 1 1 1 
[34] 1 1 0 1 1 1 1 0 0 0 1 1 1 1 0 0 0 1 1

and convert it to a factor. Identify the levels of the result, and then change the level 
labels to obtain the factor:

[1] Male Male Male Female Female Female Female Male
[10] Male Female Female Female Female Male Male Male Male Male Male Female 
      Female Female Female
[19] Female Female Female Male
[28] Female Male Male Male

Levels: Male Female
```

## Answer:
```{r}
# Create the vector
vector <- c(0, 0, 0, 1, 1, 1, 1, 0, 0, 0, 1, 1, 1, 1, 
            1, 1, 0, 1, 1, 1, 1, 0, 0, 0, 1, 1, 1, 1, 0, 0, 0, 1, 1)

# Convert it to a factor
gender_factor <- factor(vector)
levels(gender_factor)[1] <- "Male"
levels(gender_factor)[2] <- "Female"

# Print the result
print(gender_factor)

```

## Question 2:
```{latex}
  Use the `more.colors` vector, `rep()`, and `seq()` to create the vector

"red" "yellow" "blue" "yellow" "blue" "green"
"blue" "green" "magenta" "green" "magenta" "cyan"
```

## Answer:
```{r}
# R code to create the vector
more.colors <- c("red","yellow","blue","green","magenta","cyan")
more.colors[rep(seq(0 , 2) , times = 4) + rep( seq(0 : 3) , each = 3)]
```

## Question 3:
```{latex}
Convert the character vector from the preceding exercise to a factor.
What are the levels of the factor? 
Change the "blue" label to "turquoise".
```

## Answer:
```{r}
# R code to convert the character vector to a factor
factor_vector <- factor(more.colors)
cat("Labels for the factor vector: " , labels(factor_vector))
levels(factor_vector)[levels(factor_vector) == "blue"] <- "turquoise"
factor_vector
```


# Text Book problem:

## Question 4:

Use R to identify the elements of the sequence $\{2^1, 2^2, \ldots, 2^{15}\}$ that exceed the corresponding elements of the sequence $\{1^3, 2^3, \ldots, 15^3\}$.


## Answer:
```{r}
n <- 1 : 15
m1 <- 2^n # sequence 1
m2 <- n^3 # sequence 2

print("The elements in the first sequence that are greater than the last sequence:")
m1[m1 > m2]
```


## Question 5:
Evaluation of a square root is achieved using the \texttt{sqrt()} function, but a warning will be issued when the argument is negative. Consider the following code which is designed to test whether a given value is positive before checking whether the square root of the value is less than 5.

```{latex}
testValue <- 7
(testValue > 0) & (sqrt(testValue) < 5)
## [1] TRUE

testValue <- -7
if (testValue > 0) {
  result <- (sqrt(testValue) < 5)
} else {
  result <- FALSE
}
result
## [1] FALSE

Modify the code so that it continues to give the correct answer but without the warning.
```



## Answer:
```{r}
testValue <- 7
(testValue > 0) && (sqrt(testValue) < 5)

testValue <- -7
(testValue > 0) && (sqrt(testValue) < 5) # show answer no warning.

```


# Addictional Problems:

## 1:
Generate the following sequences using seq() and rep():

### a:
2, 4, 6, 8, ..., 22.

### Answer:
```{r}
seq(2 , 22 , by = 2)
```

### b: 
1, 5, 25, 125, ..., 9765625

### Answer:
```{r}
5^(0 : 10)
```

### c:
11, 12, 13, 11, 12, 13, ... 13, 11, 12, 13 (each entry appearing 7 times).

### Answer:
```{r}
rep((11 : 13) , 7)
```


### d:
1, 1, 1, 2, 2, 2, ..., 6, 6, 6

### Answer:
```{r}
rep((1 : 6) , each = 3)
```


### e:
60, 57, 54, ..., 15, 12


### Answer:
```{r}
seq(60, 12, by = -3)
```


### f:
1, 2, 5, 10, 20, 50, 100, 200, 500, 1000, 2000, 5000, 10000,20000, 50000

### Answer:
```{r}
initial <- c(1 , 2 , 5)

c(initial , initial ^ 10 , initial ^ 100 , initial ^ 1000 , initial ^ 10000)
```


## 2:
Create a vector containing each of the following sequences:

### a:
$\tan\left(\frac{\pi n}{7}\right)$, for $n = 0, 1, 2, \ldots, 10$.


### Answer:
```{r}
tan(pi * 0:10 / 7)
```

### b:
1, 4, 23, 122, 621, 3120, 15619, 78118, 390617.

### Answer:
```{r}
n <- 0 : 8

c(5^n - n)
```

### c:
3n - 3n, for \( n = 0, 1, 2, \ldots, 10 \).

### Answer:
```{r}
n <- 0 : 10

c(3^n - 3*n)
```

### d:
Compute the sequence \(5n \mod 3\) for \(n = 0, 1, 2, \ldots, 10\) (i.e., the remainder of \(5n/3\) for \(n = 0, 1, 2, \ldots, 10\)).

### Answer:
```{r}
c(5 * seq(0, 10) %% 3)
```

### e:
Let $S_n = \sum_{i=1}^{n} \frac{2i-1}{(-1)^{i+1}} = 1 - \frac{3}{1} + \frac{5}{1} - \frac{7}{1} + \ldots + \frac{2n-1}{(-1)^{n+1}}$. It can be shown that $\lim_{n \to \infty} S_n = \frac{\pi}{4}$. 

Evaluate $4S_{10}$, $4S_{100}$, and $4S_{1000}$. (Hint: use the `sum()` function).


### Answer:
```{r}

# Function to calculate Sn
Sn <- function(n) {
  sum((2*(1:n)-1) / (-1)^(1:n))
}


S_10 <- 4 * Sn(10)
S_100 <- 4 * Sn(100)
S_1000 <- 4 * Sn(1000)


c(S_10, S_100, S_1000)

```