---
title: "Assignment 1 for Stat 280"
author: 'Maharaj Teertha Deb , Student id: 40227747'
date: "2024-01-29"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1:

### Question (1) : 
Using one line of R code, calculate the interest earned on an investment of $2000, assuming an interest rate of 3% compounded annually, for terms of 1, 2, ..., 30 years.

### Answer:
```{r}
  interest_earned <- 2000 * (1 + 0.03)^ (1:30) - 2000 # Interest = P * (1 + R)^N - P
  interest_earned
```

### Question (2):
Using one line of R code, calculate the respective areas of the circles having radii 3,4,..., 100.

### Answer:
```{r}
  circle_areas <- pi * (3:100)^2
  circle_areas
```

### Question (3):
Calculate the sum $\sum_{j=0}^{n} r^j$, where $r$ has been assigned the value 1.08, and compare with $\frac{1 - r^{(n+1)}}{1 - r}$, for $n = 10, 20, 30, 40$. Repeat for $r = 1.06$.

### Answer:
```{r}
calculate_sum <- function(r, n) {
    return(sum(r^(0:n)))
}

r_power_n_without_formula <- function(r, n_values) {
    results <- sapply(n_values, calculate_sum, r = r)
    return(results)
}


r_power_n_using_formula <- function(r, n_values) {
    results <- (1 - r^(n_values + 1))/(1 - r)
    return(results)
}

r <- 1.08
n_values <- c(10, 20, 30, 40)

# for r = 1.08
print("For r = 1.08")

results_for_r1_without_formula <- r_power_n_without_formula(r, n_values)
cat("Results without formula: ", results_for_r1_without_formula, "\n")

results_for_r1_using_formula <- r_power_n_using_formula(r, n_values)
cat("Results using formula: ", results_for_r1_using_formula, "\n")

# for r = 1.06

cat("\n\n\n" , "For r = 1.06\n")
results_for_r1_without_formula <- r_power_n_without_formula(r, n_values)
cat("Results without formula: ", results_for_r1_without_formula, "\n")

results_for_r1_using_formula <- r_power_n_using_formula(r, n_values)
cat("Results using formula: ", results_for_r1_using_formula, "\n")
```


### Question (4):
Referring to the above question, utilize the quick formula to calculate 
$\sum_{j=0}^{n} r^j$ for $r = 1.08$, considering all values of $n$ in the range from 1 to 100. Save these 100 values in a vector.



### Answer:
```{r}
# Set the values
r <- 1.08
n_values <- 1:100

# Calculate the sum for each value of n
result_vector <- sapply(n_values, function(n) sum(r^(0:n)))

# Print the result vector
print(result_vector)


```



### Question (5):
Calculate the sum $\sum_{i=1}^{N} \frac{1}{i}$, and compare it with $\log(N) + 0.6$, for $N = 500, 1000, 2000, 4000, 8000$.

### Answer:
```{r}
# Define the values of n
n_values <- c(500, 1000, 2000, 4000, 8000)

# Function to calculate the sum
calculate_sum <- function(n) {
    return(sum(1 / (1:n)))
}

# Calculate sums for each value of n
sums <- sapply(n_values, calculate_sum)

# Print the sums
cat("Sums:", sums, "\n")

# Function to calculate log(n) + 0.6
logn_0.6 <- function(n) {
    return(log(n) + 0.6)
}

# Calculate log(n) + 0.6 for each value of n
log_sum <- sapply(n_values, logn_0.6)

# Print log(n) + 0.6
cat("log(n) + 0.6:", log_sum, "\n")

# Print the comparison
cat("Comparison:", log_sum - sums, "\n")
```


### Question (6):
Using \texttt{rep()} and \texttt{seq()} as needed, create the vectors:

\[
\begin{aligned}
&0 \, 0 \, 0 \, 0 \, 0 \, 1 \, 1 \, 1 \, 1 \, 1 \, 2 \, 2 \, 2 \, 2 \, 2 \, 3 \, 3 \, 3 \, 3 \, 3 \, 4 \, 4 \, 4 \, 4 \, 4 \\
&1 \, 2 \, 3 \, 4 \, 5 \, 1 \, 2 \, 3 \, 4 \, 5 \, 1 \, 2 \, 3 \, 4 \, 5 \, 1 \, 2 \, 3 \, 4 \, 5 \, 1 \, 2 \, 3 \, 4 \, 5 \\
\end{aligned}
\]


### Answer:
 To get the sequence : \[
\begin{aligned}
&0 \, 0 \, 0 \, 0 \, 0 \, 1 \, 1 \, 1 \, 1 \, 1 \, 2 \, 2 \, 2 \, 2 \, 2 \, 3 \, 3 \, 3 \, 3 \, 3 \, 4 \, 4 \, 4 \, 4 \, 4 \\
\end{aligned}
\]
```{r}
rep(seq(1 : 4) , each = 5)
```

To get the sequence : \[
\begin{aligned}
&1 \, 2 \, 3 \, 4 \, 5 \, 1 \, 2 \, 3 \, 4 \, 5 \, 1 \, 2 \, 3 \, 4 \, 5 \, 1 \, 2 \, 3 \, 4 \, 5 \, 1 \, 2 \, 3 \, 4 \, 5 \\
\end{aligned}
\]
```{r}
rep(1 : 5 , 5)
```



### Question (7):
Using \texttt{rep()} and \texttt{seq()} as needed, create the vector:

\[
1 \, 2 \, 3 \, 4 \, 5 \, 2 \, 3 \, 4 \, 5 \, 6 \, 3 \, 4 \, 5 \, 6 \, 7 \, 4 \, 5 \, 6 \, 7 \, 8 \, 5 \, 6 \, 7 \, 8 \, 9
\]


### Answer:
```{r}
  seq(1 : 5) + rep(0 : 4 , each = 5)
```






# Part 2:

### Question (1):
Using one line of R code, calculate the respective areas of the circles having radii 3,4,..., 100.
Create the following vectors in R using \texttt{seq()} and \texttt{rep()}:

(i) $1, 1.5, 2, 2.5, \ldots, 12$

(ii) $1, 8, 27, 64, \ldots, 1000$

(iii) $1, -\frac{1}{2}, \frac{1}{3}, -\frac{1}{4}, \ldots, -\frac{1}{100}$

(iv) $1, 0, 3, 0, 5, 0, 7, \ldots, 0, 49$

(v) $1, 3, 6, 10, 15, \ldots, \sum_{i=1}^{n} i, \ldots, 210$ [Look up \texttt{?cumsum}]

(vi) $1, 2, 2, 3, 3, 3, 4, 4, 4, 4, \ldots, 9, 10, 10, 10, 10, 10, 10, 10, 10, 10, 10$

### Answer:

### (i) Answer : 
```{r}
  seq(1 , 12 , by = 0.5)
```

### (ii) Answer :
```{r}
  seq(1 , 10) ^ 3 
```

### (iii) Answer:
```{r}
  n <- (1 : 100)

  seq <- 1/n * (-1)^(n)
  
  seq * (-1) 
```


### (iv) Answer:
```{r}
c((1:49) * (1:49 %% 2))
```


### (v) Answer:
```{r}
cumsum(1 : 15)
```

### (vi) Answer:
```{r}
n_values <- 1 : 10
rep(n_values , n_values)
```


