---
title: "Untitled"
author: "Maharaj Teertha Deb, 40227747"
date: "2024-03-21"
output: pdf_document
---

# Problem 1 :
The variable old Fib2 isn’t strictly necessary. Rewrite the Fibonacci while() loop with the update of Fib1 based just on the current values of Fib1 and Fib2.

## Solution:
```{r}
Fib1 <- 0
Fib2 <- 1
Fibonacci <- c(Fib1, Fib2)

while (Fib2 < 300) {
  nextFib <- Fib1 + Fib2
  Fibonacci <- c(Fibonacci, nextFib)
  Fib1 <- Fib2
  Fib2 <- nextFib
}

print(Fibonacci)
```


# Problem 2:
In fact, Fib1 and Fib2 aren’t necessary either. Rewrite the Fibonacci while() loop without using any variables except Fibonacci.

## Solution:
```{r}
Fibonacci <- c(0, 1)

while (Fibonacci[length(Fibonacci)] + Fibonacci[length(Fibonacci) - 1] < 300) {
  nextFib <- Fibonacci[length(Fibonacci)] + Fibonacci[length(Fibonacci) - 1]
  Fibonacci <- c(Fibonacci, nextFib)
}
print(Fibonacci)
```


# Problem 3:
Determine the number of Fibonacci numbers less than 1000000

## Solution:
```{r}
Fibonacci <- c(1, 1, 1)

while (max(Fibonacci) < 1000000){
  Fibonacci <- c(Fibonacci, Fibonacci[length(Fibonacci)]+
  Fibonacci[length(Fibonacci)-1])
}

sum(Fibonacci < 1000000)
```


# Problem 4:
Recall the car dealer interest rate example in Example 4.4. Use a `while ()` loop to iteratively calculate the interest rate \(i\) which satisfies the fixed-point equation 

\[
i = \frac{{1 - (1 + i)^{-20}}}{19}
\]

Use a starting guess of \(i = 0.006\). Stop the calculation when two successive values of the interest rate are less than \(0.000001\) apart. What happens when you try other starting guesses? Referring to the previous exercise, modify your code so that it also computes the number of iterations required to get two successive values of the interest rate that are less than \(0.000001\) apart.


## Solution:
```{r}
# Function to calculate the interest rate using the fixed-point equation
calculate_interest <- function(guess, tolerance = 0.000001) {
  iter <- 0
  prev_guess <- guess + 2 * tolerance # To ensure the loop starts
  
  while (abs(prev_guess - guess) >= tolerance) {
    prev_guess <- guess
    guess <- (1 - (1 + guess)^(-20)) / 19
    iter <- iter + 1
  }
  
  return(list(interest_rate = guess, iterations = iter))
}

# Initial guess for interest rate
initial_guess <- 0.006

# Calculate interest rate and number of iterations
result <- calculate_interest(initial_guess)

# Print the result
cat("Interest rate:", result$interest_rate, "\n")
cat("Number of iterations:", result$iterations, "\n")
```


# Problem 5:
Referring to the previous exercise, modify your code so that it also computes the number of iterations required to get two successive values of the interest rate that are less than 0.000001 apart.

## Solution:
```{r}
# Function to calculate the interest rate using the fixed-point equation
calculate_interest <- function(guess, tolerance = 0.000001) {
  iter <- 0
  prev_guess <- guess + 2 * tolerance # To ensure the loop starts
  
  while (abs(prev_guess - guess) >= tolerance) {
    prev_guess <- guess
    guess <- (1 - (1 + guess)^(-20)) / 19
    iter <- iter + 1
  }
  
  return(list(interest_rate = guess, iterations = iter))
}

# Initial guess for interest rate
initial_guess <- 0.006

# Calculate interest rate and number of iterations
result <- calculate_interest(initial_guess)

# Print the result
cat("Interest rate:", result$interest_rate, "\n")
cat("Number of iterations:", result$iterations, "\n")
```




# Problem 6:
Find a minimizer of the function \( f(x) = (x - 3)^4 + 7(x - 2)^2 + x \).

## Solution:
```{r}
# Define the objective function
f <- function(x) {
  return((x - 3)^4 + 7 * (x - 2)^2 + x)
}

# Use optim() function to minimize the objective function
result <- optim(par = 0, fn = f, method = "Brent", lower = -10, upper = 10)

# Print the minimizer
cat("Minimizer of the function:", result$par, "\n")
```

# Problem 7:
How many zeros does the function \( f(x) = \frac{{5x - 3}}{{x - 1}} \) have? What are they? Describe the behavior of Newton's method applied to this function if the initial guess is:

(a) 0.5

(b) 0.75

(c) 0.2

(d) 1.25


## Solution:
```{r}
# Define the function
f <- function(x) {
  return ((5*x - 3)/(x - 1))
}

# Define the derivative of the function
f_prime <- function(x) {
  return ((5*(x - 1) - (5*x - 3))/((x - 1)^2))
}

# Implement Newton's method
newton_method <- function(guess) {
  tol <- 0.0001  # tolerance for convergence
  max_iter <- 100  # maximum number of iterations
  
  x <- guess
  iter <- 0
  
  while(iter < max_iter) {
    x_new <- x - f(x)/f_prime(x)
    
    # Check if the function value is within tolerance
    if (!is.na(f(x_new)) && abs(f(x_new)) < tol) {
      return(list(zero = x_new, iterations = iter + 1))
    }
    
    x <- x_new
    iter <- iter + 1
  }
  
  # Return NaN if no convergence within max_iter
  return(list(zero = NaN, iterations = max_iter))
}

# Define initial guesses
initial_guesses <- c(0.5, 0.75, 0.2, 1.25)

# Apply Newton's method with different initial guesses
results <- lapply(initial_guesses, newton_method)

# Print results
for (i in 1:length(results)) {
  cat("Initial guess:", initial_guesses[i], "\n")
  cat("Zero found:", results[[i]]$zero, "\n")
  cat("Number of iterations:", results[[i]]$iterations, "\n\n")
}
```


## Problem 8:
How many zeros does the function \( f(x) = (x^2 - 6x + 9)e^{-x} \) have? What are they? Describe the behavior of Newton's method applied to this function if the initial guess is:
(a) \( x_0 = 3 \)
(b) \( x_0 = 3.2 \)
(c) \( x_0 = 2.99 \)
(d) \( x_0 = 3.01 \)

## Solution:
```{r}
# Define the function f(x)
f <- function(x) {
  return((x^2 - 6*x + 9) * exp(-x))
}

# Define the derivative of the function f'(x)
f_der <- function(x) {
  return((2*x - 6) * exp(-x) + (x^2 - 6*x + 9) * (-exp(-x)))
}

# Define initial guesses
initial_guesses <- c(3, 3.2, 2.99, 3.01)

# Define tolerance to stop the loop
tolerance <- 1e-3

# Run the loop for each initial guess
for (guess in initial_guesses) {
  
  # Set the initial guess
  x <- guess
  
  # Run Newton's method
  while (TRUE) {
    # Calculate next value of x
    x_next <- x - (f(x) / f_der(x))
    
    # Check if derivative of function is zero
    if (is.na(x_next)) {
      cat("Derivative of function is zero, cannot be evaluated further.\n")
      break
    }
    
    # Calculate error
    error <- abs(x_next - x)
    
    # Check if error is less than tolerance
    if (error < tolerance) {
      cat("With initial guess of", guess, "Root is", x_next, "\n")
      break
    } else {
      # Update the parameter
      x <- x_next
    }
  }
}

```




# Problem 9:
We could implement the Sieve of Eratosthenes using a while () loop.
(c) Modify this function using break to take advantage of the above result.

## Solution:
```{r}
sieve_of_eratosthenes <- function(n) {
  primes <- vector(mode = "logical", length = n + 1)
  primes[2:n] <- TRUE
  
  p <- 2
  while (TRUE) {
    # Mark multiples of p as non-prime
    primes[p * 2:n] <- FALSE
    
    # Find the next prime number
    for (i in (p + 1):n) {
      if (primes[i]) {
        p <- i
        break  # Move to the next prime
      }
    }
    
    # If there's no more prime to find, break the loop
    if (p^2 > n) {
      break
    }
  }
  
  # Return the list of prime numbers
  return(which(primes))
}

print(sieve_of_eratosthenes(3000))
```